---
title: "Washington DC Crime Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(tidyverse)
library(RColorBrewer)
```

# Washington DC Crime Analysis
<<<<<<< HEAD
## Adrian Newgard, Benjamin , Ashlynn Prisby  
=======
## Adrian Newgard, Benjamin Herschel, Ashlynn 
>>>>>>> d7f778a3d7f15b741d70021c63caca11c743e90a

As the nation's capitol, Washington, D.C. is an interesting area to study socially. In addition to being the capitol, the city is also part of a large metropolis and contains the usual economic variations therein. This analysis is focused on incidences of crime in Washington. It will consider economic, locational, and temporal factors to take a closer look at the kinds of crime that occur in one of the world's most important cities. 

## Data

Our primary data is sourced from data.gov, and was provided by the city of Washington, D.C. It contains incidences of crimes as reported to and encountered by the Washington police department, covering the year 2024. 
A secondary dataset was also used, from the American Communities Survey. It is discussed in question 1 of the 'Research Questions' section. 

### Cleaning 

Cleaning for the data consisted of removing an empty row and converting some character values to more convienent classes. During the analysis itself, some additional cleaning was required, such as using `na.rm` for some functions.

```{r, echo=FALSE}
data <- read_csv("https://raw.githubusercontent.com/1Alnew/202-Final-Project-/refs/heads/main/Crime_Incidents_in_2024.csv")
data <- data %>%  select(-OCTO_RECORD_ID)
data$DISTRICT <- as.character(data$DISTRICT)
data$WARD <- as.character(data$WARD)
```

### Variables 

```{r, echo=FALSE}
colnames(data)
```
The data contains a number of useful data points, including: 
- `OFFENSE`: The type of crime committed 
- `DISTRICT`, `WARD`, and `BLOCK`: Locational data of varying specificity. Useful for analysis at different levels. 
- `SHIFT`: The time of day that the crime took place.

## Research Questions

### Does the poverty rate of a ward effect the amount of crimes observed there?

To answer this question, the use of another dataset is required. These data come from a set published by the city of Washington DC and gathered by the American Community Survey. They contain a variety of statistics sorted by ward in the city, but for this analysis the only variable in consideration is `DP03_0128PE`, which gives percentage of a ward's population which fell below the poverty line in the previous 12 months at the time of the data's collection. 
The data reveals that, counter to what we would expect, the areas with the highest crimes observed--remember, in this dataset, they are from police encounters--are not actually in the areas with the greatest poverty. However, where the most police encounters take place makes sense when one considers what is in wards 2 and 5 particularly: high profile government buildings like the White House and the Capitol. It seems to be that police are focusing heavily on these areas to the exclusion of, or perhaps simply ignoring, more impoverished areas.

```{r, echo=FALSE}
econ <- read_csv("https://raw.githubusercontent.com/1Alnew/202-Final-Project-/refs/heads/main/ACS_5-Year_Economic_Characteristics_DC_Ward.csv") 
ward_poverty <- econ[c('NAME', 'DP03_0128PE')]
ward_poverty$NAME <- parse_number(ward_poverty$NAME)
colnames(ward_poverty)[colnames(ward_poverty) == 'NAME'] <- 'WARD'
colnames(ward_poverty)[colnames(ward_poverty) == 'DP03_0128PE'] <- '% impoverished'
ward_poverty$WARD <- as.character(ward_poverty$WARD)
wards <- data %>% drop_na(WARD) %>% count(WARD)
wards <- full_join(wards, ward_poverty, by = "WARD")
ggplot(
  wards,
  aes(x=WARD, y=n, fill=`% impoverished`)
)+geom_col()+
  scale_fill_gradient(low="#fc96ec", high = "#7c0069")+
  labs(y = "No. of Crimes", x = "Ward")
```


### Does the time of day affect where these crimes are committed? 
To get a better understanding of the crimes committed, we wanted to see if the time of day and where the crimes were located.

```{r, echo=FALSE}
library(RColorBrewer)
day_crime <- data %>%
  group_by(SHIFT,BLOCK,LONGITUDE,LATITUDE) %>%
  summarise(count =n(),.groups = "drop")

  
top_10_per_shift <- day_crime %>%
    group_by(SHIFT) %>%
    slice_max(order_by = count, n = 10, with_ties = FALSE) %>%
    arrange(SHIFT, desc(count))


top_10_per_shift <- top_10_per_shift %>%
  mutate(BLOCK = reorder(paste(BLOCK, SHIFT, sep = "___"), count))



ggplot(top_10_per_shift, aes(x = BLOCK, y = count, fill = SHIFT)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~SHIFT, scales = "free") +
  scale_x_discrete(labels = function(x) gsub("___.*", "", x)) +
  labs(
    title = "Top 10 Locations by Crime Count per Day",
    x = "Location",
    y = "Crime Count"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
By filtering out the top ten locations in comparison to what time of day they happened, it gave a better understanding when it was most common for crimes. It was seen that during the day was the busiest which is logically as it lines up with when people are at work or tourist are around. To get a deeper comparison, I filtered out overlapping locations with time of day and then created a map of the block locations within the DC area.  

```{r, echo=FALSE}
library(RColorBrewer)
top_10_per_shift$BLOCK <- sub("___.*$", "", top_10_per_shift$BLOCK)


excluded_locations <- c(
  "812 - 899 BLOCK OF BLADENSBURG ROAD NE", 
  "600 - 669 BLOCK OF PENNSYLVANIA AVENUE SE",
  "1000 - 1099 BLOCK OF 16TH STREET NW",
  "1516 - 1699 BLOCK OF BENNING ROAD NE",
  "100 - 199 BLOCK OF CARROLL STREET NW",
  "3200 - 3275 BLOCK OF M STREET NW",
  "1600 - 1699 BLOCK OF P STREET NW",
  "2600 - 2649 BLOCK OF CONNECTICUT AVENUE NW",
  "6500 - 6599 BLOCK OF GEORGIA AVENUE NW",
  "1100 - 1199 BLOCK OF U STREET NW",
  "1300 - 1399 BLOCK OF OKIE STREET NE",
  "1800 - 2299 BLOCK OF NEW YORK AVENUE NE",
  "4000 - 4121 BLOCK OF MINNESOTA AVENUE NE",
  "2100 - 2199 BLOCK OF 24TH PLACE NE",
  "1 - 7 BLOCK OF DUPONT CIRCLE NW"
)

filtered_crime <- top_10_per_shift %>%
  filter(!BLOCK %in% excluded_locations)

ggplot(filtered_crime, aes(x = BLOCK, y = count, fill = SHIFT)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~SHIFT, scales = "free") +
  scale_x_discrete(labels = function(x) gsub("___.*", "", x)) +
  labs(
    title = "Top 10 Locations by Crime Count per Day",
    x = "Location",
    y = "Crime Count"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

filtered_crime <- filtered_crime %>%
  ungroup(SHIFT)%>%
  select(-SHIFT)
filtered_crime <- filtered_crime %>%
  mutate(
    LONGITUDE = as.numeric(LONGITUDE),
    LATITUDE = as.numeric(LATITUDE)
  )


states_map <- map_data("state")

dc_map <- subset(states_map, region == "district of columbia")
ggplot() +
  geom_polygon(
    data = dc_map,
    aes(x = long, y = lat, group = group),
    fill = "gray95", color = "black"
  ) +
  geom_point(
    data = filtered_crime,
    aes(x = LONGITUDE, y = LATITUDE, color = BLOCK),  
    size = 3, alpha = 0.7
  ) +
  coord_fixed(1.3) +
  labs(
    title = "Crime Locations by Block",
    x = "Longitude",
    y = "Latitude",
    color = "Block"  
  ) +
  scale_fill_brewer(palette = "Set3") +
  theme_minimal()
  


```


It was seen that all three times of day had an overlap with block 3100-3900 Block of 14th St NW which is located in the center of DC based on the map. With some research, it was found that location is near the DC National Mall and a major Subway station which explains why it could have higher crime rates between the three times of day. Overall, there was an overlap between 7 locations with the majority being closer into the city. It was observed with outside research that those locations were near subway stations which get high traffic during all three times of day. 

### Are there known neighborhoods for a certain crime for that area? 

```{r, echo=FALSE}
library(sf)
library(RColorBrewer)
library(dplyr)
library(ggplot2)


wards_shapefile <- st_read("/Users/ashlynnprisby/Downloads/Wards_from_2022/Wards_from_2022.shp")
wards_shapefile <- wards_shapefile %>%
  mutate(WARD = as.character(WARD))


neighborhood_crimes <- data %>%
  count(OFFENSE, NEIGHBORHOOD_CLUSTER, WARD, name = "count") %>%
  filter(!is.na(OFFENSE), !is.na(NEIGHBORHOOD_CLUSTER)) %>%
  mutate(WARD = as.character(WARD))


choropleth_data_wards <- wards_shapefile %>%
  left_join(neighborhood_crimes, by = "WARD")


ggplot(data = choropleth_data_wards) +
  geom_sf(aes(fill = count), color = "white") +  # Use fill for crime count
  geom_sf_text(aes(label = WARD), color = "white", size = 3, fontface = "bold") +  # Label the wards
  scale_fill_viridis_c(option = "C") +  # Use a continuous color palette suitable for numerical data
  labs(
    title = "Crime Counts by Ward",
    fill = "Crime Count"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(size = 10))
```
```{r}
library(RColorBrewer)
Cluster_crime <- data %>%
  distinct(OFFENSE, NEIGHBORHOOD_CLUSTER, LATITUDE, LONGITUDE)
Cluster_crime <- Cluster_crime %>%
  filter(!is.na(OFFENSE), !is.na(NEIGHBORHOOD_CLUSTER))

ggplot(Cluster_crime, aes(x = LONGITUDE, y = LATITUDE, color = OFFENSE)) +
  geom_point(size = 3, alpha = 0.7) +  
  facet_wrap(~ NEIGHBORHOOD_CLUSTER) +  
  labs(
    title = "Crime Locations by Neighborhood Cluster and Offense Type",
    x = "Longitude",
    y = "Latitude",
    color = "Neighborhood Cluster"
  ) +
  scale_color_brewer(palette = "Set3") +  
  theme_minimal() + 
  theme(
    strip.text = element_text(size = 10), 
    legend.position = "right" 
  )
```


[content]

```{r, echo=FALSE}

data <- data %>%
  filter(!is.na(SHIFT), !is.na(OFFENSE), !is.na(METHOD))

data <- data %>%
  mutate(
    SHIFT = as.factor(SHIFT),
    OFFENSE = as.factor(OFFENSE),
    METHOD = as.factor(METHOD)
  )

# 1a
ggplot(data, aes(x = SHIFT)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Crime Frequency by Shift",
    x = "Shift",
    y = "Number of Crimes"
  ) +
  theme_minimal()

```

```{r, echo=FALSE}

# 1b
top_offenses <- data %>%
  count(OFFENSE, sort = TRUE) %>%
  top_n(5, n) %>%
  pull(OFFENSE)

data %>%
  filter(OFFENSE %in% top_offenses) %>%
  ggplot(aes(x = SHIFT, fill = OFFENSE)) +
  geom_bar(position = "dodge") +
  theme(axis.text.x=element_blank())+
  scale_fill_brewer(palette = "Set3")+
  labs(
    title = "Top 5 Crime Types by Shift",
    x = "Shift",
    y = "Number of Crimes",
    fill = "Offense Type"
  ) +
  theme_minimal()

```

[content]

### Question 5 

```{r, echo=FALSE}

# 2b
data %>%
  filter(OFFENSE %in% top_offenses) %>%
  ggplot(aes(x = METHOD, fill = OFFENSE)) +
  geom_bar(position = "dodge") +
  theme(axis.text.x=element_blank())+
  scale_fill_brewer(palette = "Set3")+
  facet_wrap(~ OFFENSE, scales = "free_x") +
  labs(
    title = "Common Methods Used by Top 5 Crime Types",
    x = "Method",
    y = "Number of Crimes",
    fill = "Offense"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# 2b (LOG)
data %>%
  filter(OFFENSE %in% top_offenses) %>%
  ggplot(aes(x = METHOD, fill = OFFENSE)) +
  geom_bar(position = "dodge") +
  theme(axis.text.x=element_blank())+
  scale_fill_brewer(palette = "Set3")+
  facet_wrap(~ OFFENSE, scales = "free_x") +
  scale_y_log10() +
  labs(
    title = "Common Methods Used by Top 5 Crime Types (Log Scale)",
    x = "Method",
    y = "Log10(Number of Crimes)",
    fill = "Offense"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

[content]

### How many and what crimes occur near the White House and Capitol?

In addition to providing an interesting insight into what goes on around the seat of the Federal Government, this question can make a comparison between the two that may reveal how the public interacts with them differently. 
Firstly, the White House. 
```{r, echo=FALSE}
# Filtering to make finding block which contains White House easier
penn <- data %>% filter(grepl("PENNSYLVANIA AVENUE", data$BLOCK, fixed = TRUE))
#unique(penn$BLOCK) Looking at values to find which contains the White House
penn <- penn %>% filter(BLOCK == "1500 - 1649 BLOCK OF PENNSYLVANIA AVENUE SE")
penn %>%  ggplot(
  aes(x=OFFENSE, fill=OFFENSE)
)+geom_bar()+
  theme(axis.text.x=element_blank())+
  scale_fill_brewer(palette = "Set3")+
  labs(title="Crimes on the White House Block")
```

Then, the Capitol. The data has a built-in indicator for whether a crime took place on Capitol Hill. 
```{r, echo=FALSE}
capitol <- data %>% filter(BID == "CAPITOL HILL")
dim(capitol)
capitol %>%  ggplot(
  aes(x=reorder(factor(OFFENSE), OFFENSE, length), fill=OFFENSE)
)+geom_bar()+
  scale_fill_brewer(palette = "Set3")+
  theme(axis.text.x=element_blank())+
  labs(title="Crimes on Capitol Hill",
       x = "Crime")
```

We see that there are many more crimes recorded around Capitol Hill than the White House. Total, there are 437. There a couple of possibilities for why this might be. One could be that the heavier security at the White House acts to keep more criminals away from the building than does the security at the Capitol. It could also be that there is simply more opportunity to commit crims in the Capitol Hill area.

## Conclusion

It is evident that the uniqueness of Washinton's purpose as a city leads to some unexpected occurences in the data. The presence of the seats of government in the city--the White House, the Capitol, etc.--leads to increased presence around those areas. The result is an uneven police distribution across the city in a manner different from other cities. However, D.C. suffers similar problems to other cities. 